---
title: "Maryland Poverty Level - Data Cleaning"
author: "Lebo Sango"
date: "2024-04-16"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("GGally")
#install.packages("mapproj")
#install.packages("maps")
#install.packages("readxl")


library(fpp3)
library(maps)
library(usmap)
library(usmapdata)
library(mapproj)
library(ggplot2)
library(lubridate)
library(readxl)
library("GGally")

```

## 1.1 SAIPE

```{r SAIPE, warning=FALSE}

MD_raw = read.csv("/Users/lebsan/Documents/STAT 5084 - Time Series/County Level Project/Gathering Dataset/SAIPE Masterfile.csv")

# Data Transformation

MD_transf = MD_raw %>% rename(county_pov = Poverty.Universe, FIPS = ID, 
                              poverty_pop = Number.in.Poverty) %>% 
  select(Year, FIPS, county_pov,poverty_pop, Name) %>%
  mutate(poverty_pop = as.numeric(stringr::str_remove_all(poverty_pop,",")))%>% 
  mutate(county_pov = as.numeric(stringr::str_remove_all(county_pov,",")))%>% 
  as_tsibble(index = Year, key = c(FIPS, Name)) %>% 
  na.omit() %>%
  filter( FIPS != 0)%>% 
  filter( FIPS != 24000 )# removed the zero values that correspond to the United
  #States in FIPS 


# ----------------- This code is for part I

MD_SAIPE = MD_transf%>% select(FIPS, year = Year,Name ,poverty_pop,county_pov) %>%
 filter(FIPS %in% c(24031,24033,24005,24003,24510,24027,24021, 24025, 24013))

#--------------------------

# The state of Maryland counts 24 counties
MD_transf %>% count(n_distinct(FIPS))

# The largest county by population is the 24031 or Montgomery County
MD_transf %>% arrange(desc(county_pov)) %>% 
  head()

# The 9 largest county by population 

MD_transf %>% group_by(FIPS, Name) %>%
  summarise(Largest_counties_by_pop = max(county_pov)) %>% 
  arrange(desc(Largest_counties_by_pop)) %>% filter(Year==2022) %>%
  head(9)



```

# Map of counties of the state of Maryland

```{r pressure, echo=FALSE}

# Plot the map using ggplot2

plot_usmap(data = countypop, values = "pop_2022", include = "MD",
           labels = TRUE )+
  scale_fill_continuous(
    low = "white", high = "blue", name = "Population (2022)",
    label = scales::comma
  ) 



```

# Time plot showing the level of poverty in the 9 largest counties

```{r}

MD_transf %>% select(FIPS,Year,Name ,poverty_pop) %>% 
  filter(FIPS %in% c(24031,24033,24005,24003,24510,24027,24021, 24025, 24013))  %>% 
  ggplot(aes(x= Year, y=poverty_pop, group = FIPS, color=Name))+
  geom_line()+ 
  labs(x= "Year", y= " SAIPE per County Population", 
       title = " Census of Maryland Counties' Small Area Income and Poverty, SAIPE
       ")+
  theme_light() 



```

# 1.2 County SNAP Benefits

```{r}


MD_snap = read.csv("/Users/lebsan/Documents/STAT 5084 - Time Series/County Level Project/Gathering Dataset/cntysnap (1).csv", colClasses = "character" )

# gsub removes the "X" next to year & # pivot longer aids to transpose year and 
#values to columns

snap_MD = as_tibble(MD_snap) %>% 
  pivot_longer(cols= X2021:X1989, 
                names_to = "year", 
                values_to ="SnapBenefits",
                values_drop_na = TRUE) %>% 
  mutate(year = gsub("X","",year))


# Converting year to date and merging state & county FIPS
# grepl allows to filter specifically by the string ending with

snap_MD1 = snap_MD %>% mutate(year = as.numeric(year)) %>% 
  filter(grepl(", MD", Name)) %>% 
  mutate(FIPS = paste(State.FIPS.code,County.FIPS.code ,sep = ""))%>%  
  as_tsibble(index = year, key = c(Name, FIPS) )            


# Time plot of the nine largest counties receiving SNAP benefits and converting
#Snap benefits to numeric



#------------------------------
snap_MD2 = snap_MD1 %>% select(c(FIPS,year, Name, SnapBenefits)) %>%
  filter(Name %in% c("Montgomery County, MD", "Prince George's County, MD"
                   , "Baltimore County, MD", "Anne Arundel County, MD",
                   "Baltimore city, MD", "Howard County, MD",
                   "Frederick County, MD", "Harford County, MD", 
                   "Carroll County, MD" ) ) %>%
  mutate(SnapBenefits = as.numeric(gsub("[\\,]","", SnapBenefits))) 

snap_MD2 %>% head()


# Plot of the nine Largest counties receiving SNAP

snap_MD2 %>% ggplot(aes(x=year, y=SnapBenefits, color = Name))+
  geom_line() + labs(title = " County Populaion Receiving SNAP Assistance",
                     y="SNAP per pop")+theme_linedraw()
```

#1.3 State IRS Data

```{r}

MD_tax_exempt = read_xls("/Users/lebsan/Documents/STAT 5084 - Time Series/County Level Project/Gathering Dataset/irs.xls", skip = 1)

MD_tax_exempt1 = MD_tax_exempt %>% 
  select(FIPS= `State FIPS code`,Year ,Name,
         Pop_poor_exempt = `Poor exemptions`) %>% 
   as_tsibble(index = Year, key= c(FIPS, Name)) %>% 
   filter(FIPS == 24) %>% rename(year = Year)

MD_tax_exempt1%>% head() 


# Plot of the Number of poor exemptions filed in Maryland

MD_tax_exempt1 %>% ggplot(aes(x=year, y=Pop_poor_exempt))+
  geom_line()+ 
  labs(title = "Number of Maryland Population Tax-exempted")+
  scale_y_log10()
 
```

# 1.4 Merging Data

```{r warning=FALSE}

md_merge = merge(MD_SAIPE, snap_MD2, by = c("FIPS","year"))

final_md = merge(md_merge, MD_tax_exempt1, by = c("year")) %>% 
  select(c(year,FIPS=FIPS.x, County = Name.x,State = Name, 
           SAIPE = poverty_pop, SNAP = SnapBenefits, 
           IRS_exempt_State = Pop_poor_exempt,Poverty_Universe = county_pov))

final_md %>% head()



#  Data Visualization exhibiting the relationshp between SAIPE, SNAP, IRS tax-exempt
# and poverty universe.
final_md %>% mutate(SAIPE = log(SAIPE), SNAP = log(SNAP),
                    IRS_exempt_State = log(IRS_exempt_State), 
                    Poverty_Universe = log(Poverty_Universe)) %>% 
  select(c(SAIPE,SNAP, IRS_exempt_State,
           Poverty_Universe))%>% ggpairs()





```


SAIPE and SNAP shows a high correlation coefficient and poverty universe displays a high correlation coefficient as well of  0.9. However, IRS tax-exempt has the weakest relationship with poverty universe as compared to SNAP and SAIPE. The log-transformation allows to get rid of variations in the data and capture outliers. SAIPE relationships with  SNAP and Poverty_universe exhibit sensitivity to large numbers.

